Statistical Guarantees for Local Spectral Clustering on Random Neighborhood Graphs

### 2021-09-05 jmlr/revision2/ppr_clustering_neighborhood_graph.tex
Through checking + copy-editting Appendices A and B.

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{float}
\usepackage[export]{adjustbox}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{bm}
\usepackage{mathtools}

@@

\subsection{Population normalized cut, conductance, and local spread}
Next we define three population-level functionals of $\mc{C}$---the normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and local spread $s_{\Pbb,r}(\mc{C})$---which govern the volume of the symmetric set difference $\Delta(\wh{C},\mc{C}[X])$. Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e. let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$. Also let $\deg_{\Pbb,r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\Pbb(y)$ to be the expected degree of $x$ in $G_{n,r}$. 
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

@@

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities in turn suffice to upper bound the volume of the symmetric set difference. For this reason, similar population level functionals are used by \citep{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}.

@@

Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} informally summarizes the second of our main results (formally stated in Theorem~\ref{thm:density_cluster_volume_ssd_ub}) regarding the recovery of a $\lambda$-density cluster $\mc{C}_{\lambda}$ by PPR. For reasons that we explain later in Section~\ref{sec:ppr_density_cluster}, our cluster recovery statement will actually be with respect to the $\sigma$-thickened set $\mc{C}_{\lambda,\sigma} := \{x \in \Rd: \mathrm{dist}(x,\mc{C}_{\lambda}) < \sigma\}$, for a given $\sigma > 0$. The upper bound we establish is a function of various parameters that measure the conditioning of both the density cluster $\mc{C}_{\lambda,\sigma}$ and density $f$ for recovery by PPR. We assume that $\mc{C}_{\lambda,\sigma}$ is the image of a convex set $\mc{K}$ of finite diameter $\mathrm{diam}(\mc{K}) \leq \rho < \infty$ under an Lipschitz, measure-preserving mapping $g$, with Lipschitz constant $L$. We also assume that $f$ is bounded away from $0$ and $\infty$ on $\mc{C}_{\lambda,\sigma}$,
\begin{equation*}
0 < \lambda_{\sigma} \leq f(x) \leq \Lambda_{\sigma} < \infty~~\textrm{for all $x \in \mc{C}_{\lambda,\sigma}$},
\end{equation*}
while satisfying the following low-noise condition:
\begin{equation*}
\inf_{y \in \mc{C}_{\lambda,\sigma}} f(y) - f(x) \geq  \theta \cdot \dist(x, \mc{C}_{\lambda,\sigma})^{\gamma}~~\textrm{for all $x$ such that $\dist(x,\mc{C}_{\lambda,\sigma}) \leq r$.}
\end{equation*}
(Here $\dist(x,\mc{C}) := \inf_{y \in \mc{C}} \|y - x\|$.) 
\begin{theorem}[Informal]
	\label{thm:density_cluster_volume_ssd_ub_informal}
	Let $\mc{C}_{\lambda} \subset \Rd$ be a $\lambda$-density cluster of a distribution $\Pbb$ that satisfies appropriate regularity conditions, and suppose Algorithm~\ref{alg:ppr} is well-initialized with respect to $\mc{C}_{\lambda,\sigma}$. For all $n \in \mathbb{N}$ sufficiently large, with high probability it holds that
	\begin{equation}
	\label{eqn:density_cluster_volume_ssd_ub_informal}
	\frac{\Delta(\wh{C},\mc{C}_{\lambda,\sigma}[X])}{\vol_{n,r}(\mc{C}_{\lambda,\sigma})} \leq c \cdot d^4 \cdot \frac{L^2\rho^2}{\sigma r} \cdot \frac{\Lambda_{\sigma}^2 \lambda (\lambda - \theta \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}^4} \cdot \log^2\biggl(\frac{\Lambda_{\sigma}^{2/d} L\rho}{\lambda_{\sigma}^{2/d}2r}\biggr).
	\end{equation}
\end{theorem}
Equation~\eqref{eqn:density_cluster_volume_ssd_ub_informal} reveals the separate roles played by geometry and density in the ability of PPR to recover a density cluster. The parameters $L$, $\rho$ and $\sigma$ capture whether $\mc{C}_{\lambda,\sigma}$ is well-conditioned (short and fat) or poorly-conditioned (long and thin) for recovery by PPR. Likewise, the parameters $\lambda_{\sigma}, \Lambda_{\sigma}, \gamma$ and $\theta$ measure whether $f$ is well-conditioned (approximately uniform over the density cluster, and having thin tails outside the density cluster) or poorly conditioned (vice versa). Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} tells us that if the thickened density cluster $\mc{C}_{\lambda,\sigma}$ is well-conditioned---i.e. $L^2\rho^2/(\sigma r) \approx 1$---and the density $f$ is well-conditioned near $\mc{C}_{\lambda,\sigma}$---i.e. $\Lambda_{\sigma} \approx \lambda \approx \lambda_{\sigma}$ and $\lambda - \theta r^{\gamma}/(\gamma + 1)$ is much less than $\lambda_{\sigma}$---then PPR will approximately recover $\mc{C}_{\lambda,\sigma}$.

@@

\paragraph{Spectral clustering.}
In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. In the SBM, the ability of spectral algorithms to perform clustering---or community detection---is well-understood, dating 
back to \citet{mcsherry2001} who gives conditions under which the entire
community structure can be recovered. In more recent work, \citet{rohe2011}
upper bound the fraction of nodes misclassified by a spectral algorithm for the
high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these
results to the sparse (low average degree) regime. Relatedly,
\citet{clauset08,balakrishnan2011,li2018}, analyze the misclassification rate
when the block model exhibits some hierarchical structure. The framework we
consider, in which nodes correspond to data points sampled from an underlying 
density, and edges between nodes are formed based on geometric proximity, is
quite different than the SBM, and therefore so is our analysis.

In general, the study of spectral algorithms on neighborhood graphs has been
focused on establishing asymptotic convergence of eigenvalues and eigenvectors
of certain sample objects to the eigenvalues and eigenfunctions of corresponding
limiting operators. \citet{koltchinskii2000} establish convergence of spectral
projections of the adjacency matrix to a limiting integral operator, with
similar results obtained using simplified proofs in
\citet{rosasco10}. \citet{vonluxburg2008} studies convergence of eigenvectors of
the Laplacian matrix for a neighborhood graph of fixed radius. \citet{belkin07} and
\citet{garciatrillos18} extend these results to the regime where the radius $r
\to 0$ as $n \to \infty$.

These results are of fundamental importance. However, they remain silent on the following natural question: do the spectra of these continuum operators induce a partition of the sample space which is ``good'' in some sense? \citet{shi2009,schiebinger2015,garciatrillos19,hoffmann2019} address this question, showing that spectral algorithms will recover the latent labels in certain well-conditioned nonparametric mixture models. These works are probably the most similar to our own: the conditioning of these mixture models depend on population-level functionals resembling the population-level normalized cut and conductance introduced above, and the resulting bounds on the error of spectral clustering are comparable to those we establish in Theorem~\ref{thm:volume_ssd_ub}. However, these results focus on global rather than local methods, and impose global rather than local conditions on $\Pbb$. Moreover, they do not explicitly consider recovery of density clusters, which is an important concern of our work. We comment further on the relationship between our results and these works after Theorem~\ref{thm:volume_ssd_ub}.

@@

\subsection{Organization}
We now outline the rest of the paper.
\begin{itemize}
	\item In Section~\ref{sec:ub_symmetric_set_difference}, we derive bounds on the error of PPR as a function of sample normalized cut, conductance, and local spread. We then show that under certain conditions the sample normalized cut, conductance, and local spread are close to their population-level counterparts, with high probability for sufficient number of samples. As a result, we obtain an upper bound on $\Delta(\wh{C},\mc{C}[X])/\vol_{n,r}(\mc{C}[X])$ purely in terms of these population functionals (Theorem~\ref{thm:density_cluster_volume_ssd_ub}).
	\item In Section~\ref{sec:ppr_density_cluster}, we focus on the special case where the candidate cluster $\mc{C} = \mc{C}_{\lambda}$ is a \emph{$\lambda$-density cluster}---that is, a connected component of the upper level set $\{x: f(x) \geq \lambda\}$. We derive bounds on the population normalized cut, conductance, and local spread of the density cluster, which depend on $\lambda$ as well as some other natural parameters. This leads to an upper bound on the symmetric set difference between $\wh{C}$ and the $\lambda$-density cluster. (Theorem~\ref{thm:density_cluster_volume_ssd_ub}).
	\item In Section~\ref{sec:lower_bound}, we prove a negative result: we give a hard distribution $\Pbb$ with corresponding density cluster $\mc{C}_{\lambda}$ for which the symmetric set difference between $\wh{C}$ and the $\lambda$-density cluster is provably large.
	\item In Section~\ref{sec:experiments} we empirically investigate some of our conclusions, before ending with some discussion in Section~\ref{sec:discussion}.
\end{itemize}

@@

In the main result (Theorem~\ref{thm:volume_ssd_ub}) of this section, we give a high probability upper bound on $\Delta(\wh{C}, \mc{C}[X])$, in terms of the population normalized cut $\Phi_{\Pbb,r}(\mc{C})$ and conductance $\Psi_{\Pbb,r}(\mc{C})$. We build to this theorem slowly, giving new structural results in two distinct directions. First, we build on some previous work (mentioned in the introduction) to relate $\Delta(\wh{C}, \mc{C}[X])$ to the sample normalized cut $\Phi_{n,r}(\mc{C}[X])$, conductance $\Psi_{n,r}(\mc{C}[X])$, and local spread $s_{n,r}(\mc{C}[X]) := s\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr)$. Second, we argue that when $n$ is large, each of these graph functionals can be bounded by their population-level analogues.

@@

In their most general form, the results of~\citet{zhu2013} depend on the mixing time of a lazy random walk over the induced subgraph $G[C]$. The \emph{mixing time} of a lazy random walk over a graph $G$ is
\begin{equation}
\label{eqn:mixing_time}
\tau_{\infty}(G) := \min\set{ t: \frac{{\pi}(u) - {q}_{v}^{(t)}(u)}
	{{\pi}(u)} \leq \frac{1}{4}, \; \text{for all $u,v \in V$}};
\end{equation}
here $q_v^{(t)} := e_v W^t$ is the distribution of a lazy random walk over $G$ that originates at node $v$ and runs for $t$ steps, and $\pi := \lim_{t \to \infty} q_v^{(t)}$ is the limiting distribution of $q_v^{(t)}$.
\begin{lemma}[Lemma~3.4 of \cite{zhu2013}]
	\label{lem:zhu}
	For a set $C \subseteq V$, suppose that
	\begin{equation}
	\label{eqn:zhu_condition}
	\alpha \leq \min\Bigl\{\frac{1}{45}, \frac{1}{2\tau_{\infty}(G[C])}\Bigr\},~~ \beta \leq \frac{1}{5\vol(C;G)}
	\end{equation}
	Then there exists a set $C^g \subset C$ with $\vol(C^g;G) \geq \frac{1}{2}\vol(C^g;G)$ such that for any $v \in C^g$, the sweep cut $S_{\beta,v}$ satisfies
	\begin{equation}
	\label{eqn:zhu_ub}
	\vol(S_{\beta,v} \vartriangle C;G) \leq 6\frac{\Phi(C;G)}{\alpha \beta}.
	\end{equation}
\end{lemma}
The upper bound in~\eqref{eqn:zhu_ub} does not obviously depend on the conductance $\Psi(G[C])$. However, as \cite{zhu2013} point out, letting $\pi_{\min}(G) := \min_{u \in V}\{\pi(u)\}$, it follows from Cheeger's inequality~\citep{chung1997} that 
\begin{equation}
\label{eqn:mixing_time_cheeger}
\tau_{\infty}(G) \leq \frac{\log(1/\pi_{\min}(G))}{\Psi(G)^2}.
\end{equation}
Therefore, setting (for instance) $\alpha = \frac{\Psi(G[C])^2}{2\log(1/\pi_{\min}(G))}$ and $\wh{C} = S_{\beta_0,v}$ for $\beta_0 = \frac{1}{5 \vol(C;G)}$, we obtain from~\eqref{eqn:zhu_ub} that 
\begin{equation}
\label{eqn:zhu_ub2}
\frac{\vol(C \vartriangle \wh{C};G)}{\vol(C; G)} \leq 60\frac{\Phi(C;G) \log\bigl( 1/\pi_{\min}(G[C])\bigr)}{\Psi(G[C])^2}.
\end{equation}

@@

\begin{equation}
\label{eqn:initialization}
\begin{aligned}
& \alpha \in \Bigl[(1 - 4\delta)^2, (1 - 2\delta)^2\Bigr) \cdot
 \frac{\alpha_{\Pbb,r}(\mc{C},\delta)}{2} \\
& (L,U) \subseteq \Bigl(\frac{1}{5(1 + 2\delta)},\frac{1}{5(1 + \delta)}\Bigr) \cdot 
\frac{1}{n(n - 1)\vol_{\Pbb,r}(\mc{C})}.
\end{aligned}  
\end{equation}
where
\begin{equation}
\label{eqn:alpha_initialization}
\alpha_{\Pbb,r}(\mc{C},\delta) := \frac{\ln(2)}{13} \cdot \frac{\Psi_{\Pbb,r}^2(\mc{C})}{\ln^2\Bigl(\frac{8}{(1 - 4\delta)s_{\Pbb,r}(\mc{C})}\Bigr)}.
\end{equation}

@@

By combining Lemma~\ref{lem:zhu} and Propositions~\ref{prop:pointwise_mixing_time}-\ref{prop:sample_to_population_2}, we obtain an upper bound on $\Delta(\wh{C},\mc{C}[X])$ that depends solely on the distribution $\Pbb$ and candidate cluster $\mc{C}$. To ease presentation, we introduce the \emph{condition number}, defined for a given $\mc{C} \subseteq \Rd$ and $\delta \in (0,1/4)$ as
\begin{equation}
\label{eqn:condition_number}
\kappa_{\Pbb,r}(\mc{C},\delta) := \frac{(1 + 3\delta)(1+2\delta)}{(1 - 4\delta)^2(1 - \delta)} \cdot \frac{\Phi_{\Pbb,r}(\mc{C})}{\alpha_{\Pbb,r}(\mc{C},\delta)}.
\end{equation}

@@

\clearpage
\appendix

\input{appendix.tex}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{float}
\usepackage[export]{adjustbox}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{hyperref}       % hyperlinks
\hypersetup{
	colorlinks,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue,
	bookmarks=false
}

@@

\subsection{Population normalized cut, conductance, and local spread}
Next we define three population-level functionals of $\mc{C}$---the normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and local spread $s_{\Pbb,r}(\mc{C})$---which we will use to upper bound the volume of the symmetric set difference $\Delta(\wh{C},\mc{C}[X])$ with high probability. Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e. let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$. Also let $\deg_{\Pbb,r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\Pbb(y)$ be the expected degree of $x$ in $G_{n,r}$. 
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

@@

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities can be used to upper bound the volume of the symmetric set difference. For this reason, similar population-level functionals are used by \cite{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}.

@@

Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} informally summarizes the second of our main results (formally stated in Theorem~\ref{thm:density_cluster_volume_ssd_ub}) regarding the recovery of a $\lambda$-density cluster $\mc{C}_{\lambda}$ by PPR. For reasons that we explain later in Section~\ref{sec:ppr_density_cluster}, our cluster recovery statement will actually be with respect to the $\sigma$-thickened set $\mc{C}_{\lambda,\sigma} := \{x \in \Rd: \mathrm{dist}(x,\mc{C}_{\lambda}) < \sigma\}$, for a given $\sigma > 0$. The upper bound we establish is a function of various parameters that measure the conditioning of both the density cluster $\mc{C}_{\lambda,\sigma}$ and density $f$ for recovery by PPR. We assume that $\mc{C}_{\lambda,\sigma}$ is the image of a convex set $\mc{K}$ of finite diameter $\mathrm{diam}(\mc{K}) \leq \rho < \infty$ under a Lipschitz, measure-preserving mapping $g$, with Lipschitz constant $M$. We also assume that $f$ is bounded away from $0$ and $\infty$ on $\mc{C}_{\lambda,\sigma}$,
\begin{equation*}
0 < \lambda_{\sigma} \leq f(x) \leq \Lambda_{\sigma} < \infty~~\textrm{for all $x \in \mc{C}_{\lambda,\sigma}$},
\end{equation*}
and additionally satisfies the following low-noise condition:
\begin{equation*}
\inf_{y \in \mc{C}_{\lambda,\sigma}} f(y) - f(x) \geq  \theta \cdot \dist(x, \mc{C}_{\lambda,\sigma})^{\gamma}~~\textrm{for all $x$ such that $0 < \dist(x,\mc{C}_{\lambda,\sigma}) \leq \sigma$.}
\end{equation*}
(Here $\dist(x,\mc{C}) := \inf_{y \in \mc{C}} \|y - x\|$.) 
\begin{theorem}[Informal]
	\label{thm:density_cluster_volume_ssd_ub_informal}
	Let $\mc{C}_{\lambda} \subset \Rd$ be a $\lambda$-density cluster of a distribution $\Pbb$ that satisfies appropriate regularity conditions, and suppose Algorithm~\ref{alg:ppr} is well-initialized with respect to $\mc{C}_{\lambda,\sigma}$. For all $n \in \mathbb{N}$ sufficiently large, with high probability it holds that
	\begin{equation}
	\label{eqn:density_cluster_volume_ssd_ub_informal}
	\frac{\Delta(\wh{C},\mc{C}_{\lambda,\sigma}[X])}{\vol_{n,r}(\mc{C}_{\lambda,\sigma})} \leq c \cdot d^4 \cdot \frac{L^2\rho^2}{\sigma r} \cdot \frac{\Lambda_{\sigma}^2 \lambda (\lambda - \theta \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}^4} \cdot \log^2\biggl(\frac{\Lambda_{\sigma}^{2/d} L\rho}{\lambda_{\sigma}^{2/d}2r}\biggr).
	\end{equation}
\end{theorem}
Equation~\eqref{eqn:density_cluster_volume_ssd_ub_informal} reveals the separate roles played by geometry and density in the ability of PPR to recover a density cluster. The parameters $M$, $\rho$ and $\sigma$ capture whether $\mc{C}_{\lambda,\sigma}$ is geometrically well-conditioned (short and fat) or poorly-conditioned (long and thin) for recovery by PPR. Likewise, the parameters $\lambda_{\sigma}, \Lambda_{\sigma}, \gamma$ and $\theta$ measure whether the density $f$ is well-conditioned (approximately uniform over the density cluster, and having thin tails outside the density cluster) or poorly conditioned (vice versa). Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} tells us that if the thickened density cluster $\mc{C}_{\lambda,\sigma}$ is geometrically well-conditioned---i.e. $L^2\rho^2/(\sigma r) \approx 1$---and the density $f$ is well-conditioned near $\mc{C}_{\lambda,\sigma}$---i.e. $\Lambda_{\sigma} \approx \lambda \approx \lambda_{\sigma}$ and $\lambda - \theta r^{\gamma}/(\gamma + 1)$ is much less than $\lambda_{\sigma}$---then PPR will approximately recover $\mc{C}_{\lambda,\sigma}$.

@@

\paragraph{Spectral clustering.}
In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. In the SBM, the ability of spectral algorithms to perform clustering---or community detection---is well-understood, dating 
back to \citet{mcsherry2001} who gives conditions under which the entire
community structure can be recovered. In more recent work, \citet{rohe2011}
upper bound the fraction of nodes misclassified by a spectral algorithm for the
high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these
results to the sparse (low average degree) regime. Relatedly,
\citet{clauset08,balakrishnan2011,li2018} analyze the misclassification rate
when the block model exhibits some hierarchical structure. The framework we
consider, in which nodes correspond to data points sampled from an underlying 
density, and edges between nodes are formed based on geometric proximity, is
quite different than the SBM, and therefore so is our analysis.

In general, the study of spectral algorithms on neighborhood graphs has
focused on establishing asymptotic convergence of eigenvalues and eigenvectors
of certain sample objects to the eigenvalues and eigenfunctions of corresponding
limiting operators. \citet{koltchinskii2000} establish convergence of spectral
projections of the adjacency matrix to a limiting integral operator, with
similar results obtained using simplified proofs in
\citet{rosasco10}. \citet{vonluxburg2008} studies convergence of eigenvectors of
the Laplacian matrix for a neighborhood graph of fixed radius. \citet{belkin07} and
\citet{garciatrillos18} extend these results to the regime where the radius $r
\to 0$ as $n \to \infty$.

These results are of fundamental importance. However, they remain silent on the following natural question: do the spectra of these continuum operators induce a partition of the sample space which is ``good'' in some sense? \citet{shi2009,schiebinger2015,garciatrillos19,hoffmann2019} address this question, showing that spectral algorithms will recover the latent labels in certain well-conditioned nonparametric mixture models. These works are probably the most similar to our own: the conditioning of these mixture models depends on population-level functionals resembling the population normalized cut and conductance introduced above, and the resulting bounds on the error of spectral clustering are comparable to those we establish in Theorem~\ref{thm:volume_ssd_ub}. However, these results focus on global rather than local methods, and impose global rather than local conditions on $\Pbb$. Moreover, they do not explicitly consider recovery of density clusters, which is an important concern of our work. We comment further on the relationship between our results and these works after Theorem~\ref{thm:volume_ssd_ub}.

@@

\subsection{Organization}
We now outline the rest of the paper.
\begin{itemize}
	\item In Section~\ref{sec:ub_symmetric_set_difference}, we derive bounds on the error of PPR as a function of sample normalized cut, conductance, and local spread. We then show that under certain conditions the sample normalized cut, conductance, and local spread are close to their population-level counterparts, with high probability for sufficient number of samples. As a result, we obtain an upper bound on $\Delta(\wh{C},\mc{C}[X])/\vol_{n,r}(\mc{C}[X])$ purely in terms of these population-level functionals (Theorem~\ref{thm:density_cluster_volume_ssd_ub}).
	\item In Section~\ref{sec:ppr_density_cluster}, we focus on the special case where the candidate cluster $\mc{C} = \mc{C}_{\lambda}$ is a \emph{$\lambda$-density cluster}---that is, a connected component of the upper level set $\{x: f(x) \geq \lambda\}$. We derive bounds on the population normalized cut, conductance, and local spread of the density cluster, which depend on $\lambda$ as well as some other natural parameters. This leads to an upper bound on the symmetric set difference between $\wh{C}$ and the $\lambda$-density cluster. (Theorem~\ref{thm:density_cluster_volume_ssd_ub}).
	\item In Section~\ref{sec:lower_bound}, we prove a negative result: we give a hard distribution $\Pbb$ with corresponding density cluster $\mc{C}_{\lambda}$ for which the symmetric set difference between $\wh{C}$ and the $\lambda$-density cluster is provably large.
	\item In Section~\ref{sec:experiments} we empirically investigate some of our conclusions, before ending with some discussion in Section~\ref{sec:discussion}.
\end{itemize}

@@

In the main result (Theorem~\ref{thm:volume_ssd_ub}) of this section, we give a high probability upper bound on the volume of the symmetric set difference $\Delta(\wh{C}, \mc{C}[X])$, in terms of the population normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and local spread $s_{\Pbb,r}(\mc{C})$. We build to this theorem slowly, giving new structural results in two distinct directions. First, we build on some previous work (mentioned in the introduction) to relate $\Delta(\wh{C}, \mc{C}[X])$ to the sample normalized cut $\Phi_{n,r}(\mc{C}[X])$, conductance $\Psi_{n,r}(\mc{C}[X])$, and local spread $s_{n,r}(\mc{C}[X]) := s(G_{n,r}[\mc{C}[X]])$. Second, we argue that when $n$ is large, each of these graph functionals can be bounded by their population-level analogues with high probability.

@@

In their most general form, the results of~\citet{zhu2013} depend on the mixing time of a lazy random walk over the induced subgraph $G[C]$. The \emph{mixing time} of a lazy random walk over a graph $G$ is
\begin{equation}
\label{eqn:mixing_time}
\tau_{\infty}(G) := \min\set{ t: \frac{{\pi}(u) - {q}_{v}^{(t)}(u)}
	{{\pi}(u)} \leq \frac{1}{4}, \; \text{for all $u,v \in V$}};
\end{equation}
here $q_v^{(t)} := e_v W^t$ is the distribution of a lazy random walk over $G$ initialized at node $v$ and run for $t$ steps, and $\pi := \lim_{t \to \infty} q_v^{(t)}$ is the limiting distribution of $q_v^{(t)}$.
\begin{lemma}[Lemma~3.4 of \cite{zhu2013}]
	\label{lem:zhu}
	For a set $C \subseteq V$, suppose that
	\begin{equation}
	\label{eqn:zhu_condition}
	\alpha \leq \min\Bigl\{\frac{1}{45}, \frac{1}{2\tau_{\infty}(G[C])}\Bigr\},~~ \beta \leq \frac{1}{5\vol(C;G)}.
	\end{equation}
	Then there exists a set $C^g \subset C$ with $\vol(C^g;G) \geq \frac{1}{2}\vol(C;G)$ such that for any $v \in C^g$, the sweep cut $S_{\beta,v}$ satisfies
	\begin{equation}
	\label{eqn:zhu_ub}
	\vol(S_{\beta,v} \vartriangle C;G) \leq 6\frac{\Phi(C;G)}{\alpha \beta}.
	\end{equation}
\end{lemma}
The upper bound in~\eqref{eqn:zhu_ub} does not obviously depend on the conductance $\Psi(G[C])$. However, as \cite{zhu2013} point out, letting $\pi_{\min}(G) := \min_{u \in V}\{\pi(u)\}$, it follows from Cheeger's inequality~\citep{chung1997} that 
\begin{equation}
\label{eqn:mixing_time_cheeger}
\tau_{\infty}(G) \leq \frac{\log(1/\pi_{\min}(G))}{\Psi(G)^2}.
\end{equation}
Therefore, setting (for instance) $\alpha = \frac{\Psi(G[C])^2}{2\log(1/\pi_{\min}(G))}$ and $\wh{C} = S_{\beta_0,v}$ for $\beta_0 = \frac{1}{5 \vol(C;G)}$, we obtain from~\eqref{eqn:zhu_ub} that 
\begin{equation}
\label{eqn:zhu_ub2}
\frac{\vol(C \vartriangle \wh{C};G)}{\vol(C; G)} \leq 60\frac{\Phi(C;G) \log\bigl( 1/\pi_{\min}(G[C])\bigr)}{\Psi(G[C])^2}.
\end{equation}

@@

\begin{equation}
\label{eqn:initialization}
\begin{aligned}
& \alpha \in \Bigl[(1 - 4\delta)^2, (1 - 2\delta)^2\Bigr) \cdot
 \frac{\alpha_{\Pbb,r}(\mc{C},\delta)}{2} \\
& (L,U) \subseteq \Bigl(\frac{1}{5(1 + 2\delta)},\frac{1}{5(1 + \delta)}\Bigr) \cdot 
\frac{1}{n(n - 1)\vol_{\Pbb,r}(\mc{C})}.
\end{aligned}  
\end{equation}
where
\begin{equation}
\label{eqn:alpha_initialization}
\alpha_{\Pbb,r}(\mc{C},\delta) := \frac{\ln(2)}{17} \cdot \frac{\Psi_{\Pbb,r}^2(\mc{C})}{\ln^2\Bigl(\frac{32}{(1 - 4\delta)s_{\Pbb,r}(\mc{C})}\Bigr)}.
\end{equation}

@@

By combining Lemma~\ref{lem:zhu} and Propositions~\ref{prop:pointwise_mixing_time}-\ref{prop:sample_to_population_2}, we obtain an upper bound on $\Delta(\wh{C},\mc{C}[X])$ that depends solely on the distribution $\Pbb$ and candidate cluster $\mc{C}$. To ease presentation, we introduce a \emph{condition number}, defined for a given $\mc{C} \subseteq \Rd$ and $\delta \in (0,1/4)$ as
\begin{equation}
\label{eqn:condition_number}
\kappa_{\Pbb,r}(\mc{C},\delta) := \frac{(1 + 3\delta)(1+2\delta)}{(1 - 4\delta)^2(1 - \delta)} \cdot \frac{\Phi_{\Pbb,r}(\mc{C})}{\alpha_{\Pbb,r}(\mc{C},\delta)}.
\end{equation}

@@

\clearpage

\appendix
\input{appendix.tex}
